Fractional Types

### 2019-11-13 fracGC/p.tex
organizstion of paper
figure
giant experiment to see if we can get better notation and have a smoother story

\usepackage{booktabs}  
\usepackage{subcaption} 
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{bbm}
\usepackage[greek,english]{babel}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}
\usepackage[references]{agda} 
\usepackage{newunicodechar}
\usepackage[nocenter]{qtree}
\usepackage{tikz}
\usetikzlibrary{cd}
\usetikzlibrary{quotes}

\newunicodechar{𝕌}{$\mathbb{U}$}
\newunicodechar{𝔽}{$\mathbb{F}$}
\newunicodechar{𝕋}{$\mathbb{T}$}
\newunicodechar{𝟘}{$0$}
\newunicodechar{𝟙}{$1$}
\newunicodechar{𝟚}{$2$}
\newunicodechar{⋆}{${}_*$}
\newunicodechar{ᵤ}{${}_u$}
\newunicodechar{⊚}{$\fatsemi$}

@@

@@

Instead of using the Toffoli gate as a universal primitive for
reversible classical circuits, one can leverage the full force of type
theory and category theory by expressing reversible classical
computations as \emph{isomorphisms over finite
  types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667} or
\emph{equivalences over
  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective
exposes interesting mathematical structure in reversible computations
that we will exploit to solve the ``ancilla problem'' explained next.

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
combination of the Toffoli and Hadamard gates for quantum computing
should not distract from efficiency and safety concerns. The theorems
proving universality (i) assume that temporary storage (often called
\emph{ancilla bits}) may be used~\cite{Toffoli:1980}, and (ii) that
this temporary storage is returned to its initial state before
de-allocation. Indeed if no temporary storage is allowed, the Toffoli
gate is not universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as
we demonstrate using space-time tradeoffs in Sec.~\ref{sec:examples},
the more temporary storage is allowed, the more efficient certain
computations could become. The condition requiring that the temporary
storage is only de-allocated when returned to its initial condition is
a safety condition. Violating this condition destroys the symmetry
between input and output making the circuits not reversible and, in
the quantum model, causes irreversible decoherence problems. As
reviewed in Sec.~\ref{sec:examples}, ancilla bits have a number of critical
applications and yet are poorly supported in current reversible and
quantum programming languages making them a common source of bugs.

\paragraph*{Negative Entropy.}  
According to the conventional theory of
communication~\cite{Shannon1948}, a type with $N$ values is viewed as
an abstract system that has $N$ distinguishable states where the
amount of information contained in each state is $\log{N}$. This
entropy is a measure of information which materializes itself in
memory or bandwidth requirements when storing or transmitting elements
of this type. Thus a type with 8 elements needs 3 bits of memory for
storage or 3 bits of bandwidth for communication. The logarithmic map
implies that information contained in a composite state is the sum of
the information contained in its constituents. For example, the type
$A \times B$ where $A$ has two elements and $B$ has three elements can
be thought of a composite system consisting of two independent
unrelated subsystems.  Each state of the composite system therefore
contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
the information contained in each subsystem. A \emph{fractional type}
$\frac{1}{A}$ introduces negative entropy. For example, a type with
cardinality $\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is
natural to interpret this negative entropy just like we interpret
``negative money,'' as a resource (space or bandwidth) to be repaid
(reclaimed) by some other part of the system. Indeed, we will
introduce such fractional types and use them to represent ``garbage
collection processes'' that reclaim temporary storage.

\paragraph*{Contributions.}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Trace?}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Space Denotational Semantics?}

@@

>>>>

\usepackage{booktabs}  
\usepackage{subcaption} 
\usepackage{amssymb}
\usepackage{stmaryrd}
\usepackage{bbm}
\usepackage[greek,english]{babel}
\usepackage{ucs}
\usepackage[utf8x]{inputenc}
\usepackage{autofe}
\usepackage[references]{agda} 
\usepackage{newunicodechar}
\usepackage[nocenter]{qtree}
\usepackage{tikz}
\usepackage{bbold}
\usetikzlibrary{cd}
\usetikzlibrary{quotes}

\newunicodechar{𝕌}{$\mathbb{U}$}
\newunicodechar{𝔽}{$\mathbb{F}$}
\newunicodechar{𝕋}{$\mathbb{T}$}
\newunicodechar{𝟘}{$\mathbb{0}$}
\newunicodechar{𝟙}{$\mathbb{1}$}
\newunicodechar{𝟚}{$\mathbb{2}$}
\newunicodechar{⋆}{${}_*$}
\newunicodechar{ᵤ}{${}_u$}
\newunicodechar{⊚}{$\fatsemi$}

@@

% 12 pages

@@

Instead of using the Toffoli gate as a universal primitive for
reversible classical circuits, one can leverage the full force of type
theory and category theory by expressing reversible classical
computations as \emph{isomorphisms over finite
  types}~\cite{Fiore:2004,James:2012:IE:2103656.2103667} or
\emph{equivalences over
  groupoids}~\cite{DBLP:conf/esop/CaretteS16}. This perspective
exposes a rich mathematical structure in reversible computations
that we will exploit to solve the ``ancilla problem'' explained next.

\paragraph*{Temporary Storage using Ancilla Bits.} The universality of
the Toffoli gate for classical reversible computing and of the
universality of the combination of the Toffoli and Hadamard gates for
quantum computing should not distract from efficiency and safety
concerns. The theorems proving universality (i) assume that temporary
storage (often called \emph{ancilla bits}) may be
used~\cite{Toffoli:1980}, and (ii) that this temporary storage is
returned to its initial state before de-allocation. Indeed if no
temporary storage is allowed, the Toffoli gate is not
universal~\cite{DBLP:conf/innovations/AaronsonGS17} and as we
demonstrate in Sec.~\ref{sec:examples}, the more temporary storage is
allowed, the more efficient certain computations could become. The
condition requiring that the temporary storage is only de-allocated
when returned to its initial condition is a safety
condition. Violating this condition destroys the symmetry between
input and output making the circuits not reversible and, in the
quantum model, causes decoherence that may destroy the quantum
state. As reviewed in Sec.~\ref{sec:examples}, ancilla bits have a
number of critical applications and yet are poorly supported in
current reversible and quantum programming languages making them a
common source of bugs.

\paragraph*{Negative Entropy.}  
According to the conventional theory of
communication~\cite{Shannon1948}, a type with $N$ values is viewed as
an abstract system that has $N$ distinguishable states where the
amount of information contained in each state is $\log{N}$. This
entropy is a measure of information which materializes itself in
memory or bandwidth requirements when storing or transmitting elements
of this type. Thus a type with 8 elements needs 3 bits of memory for
storage or 3 bits of bandwidth for communication. The logarithmic map
implies that information contained in a composite state is the sum of
the information contained in its constituents. For example, the type
$A \times B$ where $A$ has two elements and $B$ has three elements can
be thought of a composite system consisting of two independent
unrelated subsystems.  Each state of the composite system therefore
contains $\log{(2*3)} = \log{2} + \log{3}$ bits which is the sum of
the information contained in each subsystem. If we could imagine a
\emph{fractional type} $\frac{1}{A}$, this type would introduce
negative entropy. For example, a type with ``cardinality''
$\frac{1}{8}$ has entropy $\log{\frac{1}{8}} = -3$. It is natural to
interpret this negative entropy just like we interpret ``negative
money,'' as a resource (space or bandwidth) to be repaid (reclaimed)
by some other part of the system. Indeed, we will introduce such
fractional types and use them to represent ``garbage collection
processes'' that reclaim temporary storage. Just like in the case of
``negative money'' It is critical that these fractional types be
linearly used, i.e., they cannot be duplicated or erased: this
property is however automatically enforced in a reversible
computational model!

\paragraph*{Contributions.} The paper presents a novel concept, that
is \emph{fractional types} etc.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Space Operational Semantics}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Trace}
